Using statistics to guide response to the HIV epidemic

### 2023-05-17 src/docs_17-05-23-hi-med/17-05-23-hi-med.Rmd
Almost finished with slides #144

1. Background on small-area estimation of HIV indicators
2. Estimating HIV risk group proportions
3. Is this something you might want to work on?

# 1. Background on small-area estimation of HIV indicators

# 2. Estimating HIV risk group proportions

## Background

@@

* In sub-Saharan Africa, adolescent girls and young women (AGYW) aged 15-29 are 28% of the population but 44% of new infections
* This disparity is because of:
  1. Younger age at first sex
  2. Age patterns of sexual mixing
  3. Structural vulnerabilities and power imbalances
  4. Increased susceptibility to HIV infection

@@

```{r echo=FALSE, out.width = "125px", fig.cap="Tweet from UNAIDS"}
knitr::include_graphics("tweet.png")
```

@@

## Prevention packages

* There are some prevention options ("core package") cheap enough they should be provided to all AGYW 
* Other options ("intensified interventions") are more costly, and there are insufficient resources to offer to all AGYW

\begin{center}
\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
$\implies$ How to prioritise interventions to have the greatest impact?
\end{tcolorbox}
\end{center}

## Stratified prevention

* Global AIDS strategy 2021-2026 proposed stratifying HIV prevention for AGYW based on
  1. Population-level HIV incidence
  2. Individual-level sexual risk behaviour
* These are the two most proximal drivers of sexual transmission

```{r echo=FALSE, out.width = "125px", fig.cap="Global AIDS strategy"}
knitr::include_graphics("global-aids-strategy.png")
```

# 3. Is this something you might want to work on?

@@

* Maths undergraduate $\to$ statistics masters $\to$ statistics PhD
* Interested in impactful applications of statistics
* Limited domain knowledge in any given area! Need to collaborate!

## Many skills are required

* Mathematics and statistics
* Computer science and software engineering
* Epidemiology and public health
* Health economics and behavioural science
* Implementation science
* ...

@@

## Bibliography {.allowframebreaks}

>>>>

1. Precision public health for HIV
2. Estimating risk group proportions
3. Some broader thoughts on careers

# 1. Precision public health for HIV

## What is precision public health

* In medicine, we want to deliver the right intervention, at the right time
* \textcolor{hilit}{Precision medicine} seeks to do that at the individual-level
* \textcolor{hilit}{Precision public health} seeks to do it at the population-level

##

@@

> Recognizing that **"one size does not fit all"**, the Strategy prioritizes tailoring of differentiated service packages and service delivery approaches to the **unique needs of people, communities and locations**, using granular data to focus programmes most effectively.

@@

```{r echo=FALSE, out.width = "125px", fig.cap="A key feature of the HIV response, both historically and into the future."}
knitr::include_graphics("global-aids-strategy.png")
```

@@

## How can statistics help?

* \textcolor{hilit}{Granular estimates} of relevant quantities are required
* Disaggregation by \textcolor{hilit}{geography, sex, age, and behaviour}

\begin{center}
\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]

\begin{itemize}
\item HIV prevalence: proportion of people who are infected
\item HIV incidence: rate of new infections
\item Antiretroviral therapy (ART) coverage: proportion of people living with HIV who take ART
\end{itemize}

\end{tcolorbox}
\end{center}

##

```{r echo=FALSE, out.width = "300px", fig.cap="Example population health goals."}
knitr::include_graphics("fast-track.png")
```

# 2. Estimating risk group proportions

##

```{r warning=FALSE, echo=FALSE}
cbpalette <- c("#56B4E9", "#009E73", "#E69F00", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")

data.frame(
  x = c("28% of the population", "44% of new infections"),
  y = c(0.28, 0.44),
  level = c(2, 1)
) %>%
  ggplot(aes(x = level, y = y, fill = x)) +
    geom_col() +
    coord_flip() +
    geom_text(aes(label = x, y = 0.1), colour = "white", position = "dodge", size = 5) +
    scale_y_continuous(labels = scales::percent) +
    scale_fill_manual(values = cbpalette) +
    labs(x = "", y = "", title = "Adolescent girls and young women (AGYW) 15-29 are", caption = "(Across 13 AGYW priority countries in sub-Saharan Africa)") +
    guides(fill = "none") +
    theme_minimal() +
    theme(
      axis.title.y = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank()
    ) 
```

## Why?

1. Younger age at first sex
2. Age patterns of sexual mixing
3. Structural vulnerabilities and power imbalances
4. Increased susceptibility to HIV infection

##

```{r echo=FALSE, out.width = "250px", fig.cap="Convergence of marginal utility with equity and human rights."}
knitr::include_graphics("rec5.png")
```

##

```{r echo=FALSE}
data.frame(
  price = c(20, 40, 50, 80, 70, 70, 50),
  effectiveness = c(80, 50, 60, 60, 50, 30, 40),
  name = c("Condoms and lube", "STI screening", "Male partner testing", "Access to PrEP", "Access to PEP", "Cash transfers and incentives", "Keep girls in school"),
  uncertainty = factor(c("More sure", rep("Less sure", times = 6)))
) %>%
  ggplot(aes(x = price, y = effectiveness, label = name, col = uncertainty)) +
    geom_text() +
    labs(x = "Price", y = "Real world effectiveness", title = bquote(bold("Insufficient resources") ~ "to provide all interventions to everyone"), caption = "Positions on plot depend on setting and should not be interpreted too precisely.\n Some interventions do not only have the effect of reducing HIV incidence.", col = "") +
    scale_color_manual(values = c("black", "#CC79A7")) +
    scale_x_continuous(breaks = c(10, 90), labels = c("Cheap", "Expensive"), limits = c(0, 100)) +
    scale_y_continuous(breaks = c(10, 90), labels = c("Works poorly", "Works well"), limits = c(0, 100)) +
    theme_minimal() +
    theme(
      panel.border = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.line = element_line(colour = "grey92", linewidth = 0.5, arrow = arrow(length = unit(5, "pt"), type = "open"))
    )
```

## How to prioritise interventions?

* The \textcolor{hilit}{most proximal drivers} of transmission are
$$
\mathbb{P}(\text{transmission}) \propto \text{sexual partnerships} \times \mathbb{P}(\text{each partner can transmit})
$$

##

```{r echo=FALSE}
x <- y <- c(1:4)

expand.grid(x, y) %>%
  mutate(risk = ifelse(Var2 > 1, Var1 * Var2, 0)) %>%
  ggplot(aes(x = Var1, y = Var2)) +
  geom_tile(aes(fill = risk)) +
  scale_fill_gradientn(
    colours = c("white", "#E69F00"),
    breaks = c(2, 14),
    labels = c("Less risk", "More risk") 
  ) +
  scale_x_continuous(breaks = c(1:4), labels = c("Low: <0.3%", "Moderate: 0.3-1%", "High: 1-3%", "Very high: >3%")) +
  scale_y_continuous(breaks = c(1:4), labels = c("None", "Low", "High", "Very\n high")) +
  labs(x = "Population-level HIV incidence", y = "Individual-level risk behaviour", title = bquote("Important we consider" ~ bold("both") ~ "population setting and individual behaviour"), fill = "") +
  theme_minimal() +
  theme(
    panel.border = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line = element_line(colour = "grey92", linewidth = 0.5, arrow = arrow(length = unit(5, "pt"), type = "open"))
  )
```

## Individual-level behavioural risk

| Level       | Behavioural risk group             | Risk ratio |
|-------------|------------------------------------|------------|
| None        | Not sexually active                | 0          |
| Low         | One cohabiting partner             | 1          |
| High        | Multiple or non-regular partner(s) | 1.72       |
| Very High   | Female sex workers                 | 13         |

## What have we done

* Goal: enable \textcolor{hilit}{prioritisation} of HIV programming for AGYW
* Audience: programme managers, service providers and policy makers 
* Plan: estimate district-age-behavioural risk group proportions, HIV prevalences and HIV incidences across 13 high priority countries in sub-Saharan Africa
* Method: a Bayesian spatio-temporal model of survey data

##

```{r echo=FALSE, out.width="350px", fig.cap = "A total of 46 surveys: importantance of good data processing code!"}
knitr::include_graphics("depends/available-surveys.png")
```

##

```{r echo=FALSE, warning=FALSE, out.width="350px", fig.cap = "Example of model structure: we assume that neighbouring districts are similar."}
areas <- sf::read_sf("depends/zwe_areas.geojson")

areas %>%
  filter(area_level == 2) %>%
  arealutils::plot_graph(add_geography = TRUE)
```

##

```{r echo=FALSE, out.width="80%", fig.cap = "Direct survey estimates are too noisy to work with due to low sample sizes, but our modelled estimates don't have the same issue."}
knitr::include_graphics("depends/model-direct-benefits.png")
```

##

```{r echo=FALSE, out.width="80%", fig.cap = "We found a prominent geographic discontinuity between eastern and southen Africa in risk group membership for women 20-29."}
knitr::include_graphics("depends/aids-abstract-A.pdf")
```

##

```{r echo=FALSE, out.width="80%", fig.cap = "Supposing you prioritised those in the highest incidence strata, what percent of new infections could you find per percent of the population reached?"}
knitr::include_graphics("depends/infections-reached.pdf")
```

## Where might this work go

* Extension to adults 15-49 and further countries
* Generation of estimates led by country teams as a part of a yearly process: data ownership, data use, and data quality

##

```{r echo=FALSE, out.width = "300px", fig.align='center', fig.cap = "Example of user interface for estimates generation."}
knitr::include_graphics("naomi_user.png")
```

# 3. Some broader thoughts on careers

##

* Upside: could \textcolor{hilit}{influence health outcomes for many people}
* Downside: part of a (more?) complex theory of change

\begin{align*}
\mathbb{P}(\text{impact}) = &\mathbb{P}(\text{relevant people interact with your work}) \\
&\times \mathbb{P}(\text{what you've done is useful}) \\
&\times \mathbb{P}(\text{message correctly conveyed}) \times \cdots \\
&\times \mathbb{P}(\text{intervention technically works}) \\
&\times \mathbb{P}(\text{intervention correctly implemented}) \times \cdots
\end{align*}

* How does this balance out?

@@

* Maths undergraduate $\to$ statistics masters $\to$ statistics (and machine learning, sort of) PhD
* Involved with Effective Altruism during my masters, and since then I've been interested in impactful applications of statistics

## Applied methodology

* Being good (perhaps not great) at a useful enough collection of things
* Feeling like no one care about the application at methods conferences, and no one cares about the methods at application conferences

## Team science

* A lot of disciplines required: Epidemiology, public health, health economics, behavioural science, implementation science, mathematics, statistics, computer science, software engineering
* Academia pushes against this by wanting every contribution to be marginally interesting in each field

##

```{r echo=FALSE, out.width = "300px", fig.align='center', fig.cap = "Alternative structures for science."}
knitr::include_graphics("fro.png")
```

@@

##

```{r echo=FALSE, out.width = "300px", fig.align='center', fig.cap = "For more (informal!) take a look at the blog post on my website`"}
knitr::include_graphics("blog-post-updated.png")
```

##

```{r echo=FALSE, out.width = "300px", fig.align='center', fig.cap = "For more (formal!) take a look at the paper."}
knitr::include_graphics("plos-gph.png")
```
